---
title: "P8105_HW2_sw4257"
output:  github_document
---

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1
### Import and tidy NYC subway data
```{r, message = FALSE, warning = FALSE}
subway_df <- read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
  na = c("", "NA", ".")) |>
  janitor::clean_names() |>
   select(line, station_name, station_latitude, station_longitude, 
          starts_with("route"), entry, vending, entrance_type, ada) |>
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))
```

The cleaned subway dataset contains information about subway lines, station names and locations, routes served, entrance and vending characteristics, entrance types, and ADA compliance. The cleaning process included cleaning the variable names, retaining the requested variables, and converting `entry` from a character variable to a logical variable. The resulting dataset contains ``r nrow(subway_df)`` rows and ``r ncol(subway_df)`` columns. The data are not completely tidy because the subway routes serving each station are spread across several columns.

### Summarize subway stations and entrances
```{r}
distinct_station_count = subway_df |>
  distinct(line, station_name) |>
  nrow()

ada_station_count = subway_df |>
  filter(ada) |>
  distinct(line, station_name) |>
  nrow()

entry_proportion =
  subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean(na.rm = TRUE)
```

There are ``r distinct_station_count`` distinct stations, ``r ada_station_count`` of them are ADA compliant. Among station entrances and exits without vending, ``r round(entry_proportion * 100, 1)``% allow entrance.

### Reformat route data and summarize A train stations
```{r}
subway_long =
  subway_df |>
  mutate(
    across(route1:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE)

a_train_station_count =
  subway_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

a_train_ada_count =
  subway_long |>
  filter(route_name == "A", ada) |>
  distinct(line, station_name) |>
  nrow()
```

The A train serves ``r a_train_station_count`` distinct stations and ``r a_train_ada_count`` of them are ADA compliant.


## Problem 2
### Import and tidy Mr. Trash Wheel data
```{r}
mr_trash_wheel =
  read_excel("data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N709") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(sports_balls = as.integer(round(sports_balls)),
         year = as.integer(year),
         wheel = "Mr. Trash Wheel")
```

### Import and tidy Professor Trash Wheel and Gwynnda
```{r}
professor_trash_wheel =
  read_excel("data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M134") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(year = as.integer(year),
         wheel = "Professor Trash Wheel")

gwynnda_falls_trash_wheel =
  read_excel("data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L351") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(year = as.integer(year),
         wheel = "Gwynnda Falls Trash Wheel")
```

### Combine the three datasets into a single tidy dataset
```{r}
trash_wheel_df = bind_rows(
  mr_trash_wheel,
  professor_trash_wheel,
  gwynnda_falls_trash_wheel) |>
  relocate(wheel)
```

```{r}
professor_total_weight = trash_wheel_df |>
  filter(wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum(na.rm = TRUE)

gwynnda_cigarette_butts = trash_wheel_df |>
  filter(wheel == "Gwynnda Falls Trash Wheel",
         str_trim(month) == "June",
         year == 2022) |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)
```

The combined Trash Wheel dataset contains ``r nrow(trash_wheel_df)`` observations and ``r ncol(trash_wheel_df)`` variables. Key variables include the `Trash Wheel name`, `dumpster number`, `collection date`, `trash weight`, `trash volume`, and counts of different types of trash, such as `plastic bottles` and `cigarette butts`. For the available data, Professor Trash Wheel collected a total of ``r round(professor_total_weight, 2)`` tons of trash, and Gwynnda Trash Wheel collected ``r format(gwynnda_cigarette_butts, big.mark = ",", scientific = FALSE)`` cigarette butts in June 2022.


## Problem 3
### Import and tidy baseline demographic data
```{r, message = FALSE}
baseline = 
  read_csv("data/data_mci/MCI_baseline.csv", skip = 1, na = ".") |>
  janitor::clean_names() |>
  mutate(sex = factor(sex, 
                      levels = c(0, 1), 
                      labels = c("Female", "Male")),
         apoe4 = factor(apoe4, 
                        levels = c(0, 1), 
                        labels = c("APOE4 non-carrier", "APOE4 carrier"))) |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)
```

### Summarize participant characteristics
```{r}
participants_recruited = nrow(baseline)

participants_developed_mci = sum(!is.na(baseline$age_at_onset))

average_baseline_age = mean(baseline$current_age)

women_apoe4_proportion =
  baseline |>
  filter(sex == "Female") |>
  summarize(proportion = mean(apoe4 == "APOE4 carrier")) |>
  pull(proportion)
```

The baseline dataset contains demographic and clinical information, including `participant ID`, `baseline age`, `sex`, `education`, `APOE4 carrier status`, and `age at MCI onset`. During the import process, the first descriptive row is skipped, `"."` is treated as the missing value, cleaned the variable names, converted sex and APOE4 status to factors, and removed participants who had developed MCI by baseline. The resulting dataset contains ``r participants_recruited`` participants, ``r participants_developed_mci`` of them developed MCI during follow-up. The average baseline age was ``r round(average_baseline_age, 2)`` years, and ``r round(women_apoe4_proportion * 100, 1)``% of the women were APOE4 carriers.

### Import and tidy longitudinal biomarker data
```{r, message = FALSE}
amyloid =
  read_csv("data/data_mci/mci_amyloid.csv", skip = 1, na = c("NA", "Na")) |>
   janitor::clean_names() |>
  rename(id = study_id) |>
  pivot_longer(baseline:time_8,
               names_to = "time",
               values_to = "amyloid_ratio") |>
  mutate(time = parse_number(str_replace(time, "baseline", "0")))
```
The longitudinal amyloid dataset was imported by skipping its first descriptive row and treating both `"NA"` and `"Na"` as missing values. The variable names are cleaned, renamed the participant identifier to match the baseline dataset, converted the five measurement columns into long format, and represented time as years since baseline. The resulting dataset contains ``r n_distinct(amyloid$id)`` participants and ``r nrow(amyloid)`` participant-visit records across five possible measurement times. The key variables are `id`, `time`, and `amyloid_ratio`, which represent participant ID, years since baseline, and the amyloid $\beta$ 42/40 ratio respectively. While there are ``r sum(is.na(amyloid$amyloid_ratio))`` scheduled measurements have missing biomarker values.

### Compare participants across datasets
```{r}
baseline_only = baseline |>
  distinct(id) |>
  anti_join(amyloid |> distinct(id), by = "id") 
  

amyloid_only = amyloid |>
  distinct(id) |>
  anti_join(baseline |> distinct(id), by = "id")
```

There were ``r nrow(baseline_only)`` participants who appeared only in the baseline dataset and ``r nrow(amyloid_only)`` who appeared only in the amyloid dataset. Some participants in the amyloid-only group were excluded from the cleaned baseline dataset because their recorded MCI onset age was less than or equal to their baseline age, which indicating that they had already developed MCI by the study baseline and therefore did not meet the inclusion criteria.

### Combine and export the demographic and biomarker data
```{r}
mci_combined = inner_join(baseline, amyloid, by = "id") |>
  arrange(id, time)

write_csv(mci_combined, "data/data_mci/mci_combined.csv")
```
The combined dataset contains ``r n_distinct(mci_combined$id)`` participants, ``r nrow(mci_combined)`` participant-visit records, and ``r ncol(mci_combined)`` variables. Each row represents one participant at one scheduled biomarker measurement time and includes baseline demographic information, APOE4 carrier status, MCI onset information, time since baseline, and the amyloid $\beta$ 42/40 ratio.
