---
title: "P8105_HW2_sw4257"
output:  github_document
---

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1
```{r, message = FALSE}
subway_df <- read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
  na = c("", "NA", ".")) |>
  janitor::clean_names() |>
   select(line, station_name, station_latitude, station_longitude, 
          starts_with("route"), entry, vending, entrance_type, ada) |>
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))
```

The cleaned subway dataset contains information about subway lines, station names and locations, routes served, entrance and vending characteristics, entrance types, and ADA compliance. The cleaning process included cleaning the variable names, retaining the requested variables, and converting `entry` from a character variable to a logical variable. The resulting dataset contains ``r nrow(subway_df)`` rows and ``r ncol(subway_df)`` columns. The data are not completely tidy because the subway routes serving each station are spread across several columns.

```{r}
distinct_station_count = subway_df |>
  distinct(line, station_name) |>
  nrow()

ada_station_count = subway_df |>
  filter(ada) |>
  distinct(line, station_name) |>
  nrow()

entry_proportion =
  subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean(na.rm = TRUE)
```

There are ``r distinct_station_count`` distinct stations, ``r ada_station_count`` of them are ADA compliant. Among station entrances and exits without vending, ``r round(entry_proportion * 100, 1)``% allow entrance.

```{r}
subway_long =
  subway_df |>
  mutate(
    across(route1:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE)

a_train_station_count =
  subway_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

a_train_ada_count =
  subway_long |>
  filter(route_name == "A", ada) |>
  distinct(line, station_name) |>
  nrow()
```

The A train serves ``r a_train_station_count`` distinct stations, of which ``r a_train_ada_count`` are ADA compliant.


## Problem 2
### Import and tidy Mr. Trash Wheel data
```{r}
mr_trash_wheel =
  read_excel("data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N709") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(sports_balls = as.integer(round(sports_balls)),
         year = as.integer(year),
         wheel = "Mr. Trash Wheel")
```

### Import and tidy Professor Trash Wheel and Gwynnda
```{r}
professor_trash_wheel =
  read_excel("data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M134") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(year = as.integer(year),
         wheel = "Professor Trash Wheel")

gwynnda_falls_trash_wheel =
  read_excel("data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L351") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(year = as.integer(year),
         wheel = "Gwynnda Falls Trash Wheel")
```

### Combine the three datasets into a single tidy dataset
```{r}
trash_wheel_df = bind_rows(
  mr_trash_wheel,
  professor_trash_wheel,
  gwynnda_falls_trash_wheel) |>
  relocate(wheel)
```

```{r}
professor_total_weight = trash_wheel_df |>
  filter(wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum(na.rm = TRUE)

gwynnda_cigarette_butts = trash_wheel_df |>
  filter(wheel == "Gwynnda Falls Trash Wheel",
         str_trim(month) == "June",
         year == 2022) |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)
```

The combined Trash Wheel dataset contains ``r nrow(trash_wheel_df)`` observations and ``r ncol(trash_wheel_df)`` variables. Key variables include the `Trash Wheel name`, `dumpster number`, `collection date`, `trash weight`, `trash volume`, and counts of different types of trash, such as `plastic bottles` and `cigarette butts`. For the available data, Professor Trash Wheel collected a total of ``r round(professor_total_weight, 2)`` tons of trash, and Gwynnda Trash Wheel collected ``r format(gwynnda_cigarette_butts, big.mark = ",", scientific = FALSE)`` cigarette butts in June 2022.
