---
title: "P8105_HW2_sw4257"
output:  github_document
---

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1
```{r, message = FALSE}
subway_df <- read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
  na = c("", "NA", ".")) |>
  janitor::clean_names() |>
   select(line, station_name, station_latitude, station_longitude, 
          starts_with("route"), entry, vending, entrance_type, ada) |>
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))
```

The cleaned subway dataset contains information about subway lines, station names and locations, routes served, entrance and vending characteristics, entrance types, and ADA compliance. The cleaning process included cleaning the variable names, retaining the requested variables, and converting `entry` from a character variable to a logical variable. The resulting dataset contains ``r nrow(subway_df)`` rows and ``r ncol(subway_df)`` columns. The data are not completely tidy because the subway routes serving each station are spread across several columns.

```{r}
distinct_station_count = subway_df |>
  distinct(line, station_name) |>
  nrow()

ada_station_count = subway_df |>
  filter(ada) |>
  distinct(line, station_name) |>
  nrow()

entry_proportion =
  subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean(na.rm = TRUE)
```

There are ``r distinct_station_count`` distinct stations, ``r ada_station_count`` of them are ADA compliant. Among station entrances and exits without vending, ``r round(entry_proportion * 100, 1)``% allow entrance.

```{r}
subway_long =
  subway_df |>
  mutate(
    across(route1:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE)

a_train_station_count =
  subway_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

a_train_ada_count =
  subway_long |>
  filter(route_name == "A", ada) |>
  distinct(line, station_name) |>
  nrow()
```

The A train serves ``r a_train_station_count`` distinct stations, of which ``r a_train_ada_count`` are ADA compliant.


## Problem 2
Import the Mr. Trash Wheel data
```{r}
mr_trash_wheel =
  read_excel("data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N709") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(sports_balls = as.integer(round(sports_balls)))
```

